The Evolution of Phenotypically Invariant Genetic Networks

### 2021-08-10 review-responses2.tex

\reply{
This is correct. As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids. We rely on results from quantitative genetics to estimate the rate at which systems drift through the neutral space and thus to estimate the effects of gene flow. Without further modeling we do not yet know how a quantitative genetics model incorporating system theory would differ, if it all, from these classic results. We are planning to submit a paper where we further develop the removed quantitative genetics model and may be able to address this interesting question there.

We also briefly mention that gene flow will likely complicate matters in the discussion \revref.
}

\begin{point}{Orr's Snowball}
The authors state that rapid emergence of incompatibility is possible in their model. How would the accumulation of incompatibility here compare with Orr’s ``snownball'' model, and how is this expected to depend on the complexity of the underlying networks?
\end{point}

\reply{
    Good question:
    on face value, the similarity to the snowball depends entirely on the relationship between
    ``distance to optimum'' and fitness: since distance to optimum increases linearly in F2s,
    if this relation is quadratic then our model will also show a quadratic increase in incompatibility.
    However, this seems somewhat arbitrary.
    It is more interesting to consider if our model displays the same underlying behavior that leads to the snowball,
    in which each pair of lineage-specific substitutions contributes to incompatibility.
    This may be an empirical question (that could be examined in simulation),
    but at this level of analysis it seems the answer is ``no'':
    Consider two lineages moving away from each other along the neutral manifold.
    Each substitution moves them further apart, but the rate of increase of the distance between them
    is linear, and each subsequent substitution does not increase their rate of separation.
    We considered adding this point to the paper,
    but decided that it is sufficiently confusing that it would not add much.
}

@@

We briefly mention complications to the model in the discussion \revref.
}

\reviewersection{1}
\begin{quote}
  I am basically enthusiastic about this paper, which has been influential for some time on bioRxiv. The central sections, especially e.g. \hyperlink{rev1:1}{lines 127-219}, are very clear and novel. I do think, though that the work could be better placed in context.
\end{quote}

Thank you. We were also excited that the paper had many readers on the bioRxiv.

\begin{point}{Empirical evidence for the (minor?) role of drift in speciation:}
%1. Empirical evidence for the (minor?) role of drift in speciation:
%
    \revref~discusses previous discussion about the role of drift in speciation, but they focus on theory. I think that some strong arguments for the primacy of selection in speciation have been empirical (bottlenecking Drosophila populations and testing for RI, MK-tests on ``speciation genes'' etc.). Some of the earlier work was summarized in the last chapter of Coyne and Orr 2004. \llname{rp11}~seems to ignore a lot of this work, and so I would rephrase.
\end{point}

@@

\reply{This is an important point. We have added references to the experimental speciation literature \revref. 
}

\begin{point}{Additive quantitative trait models}
%2. Additive quantitative trait models
%
The previous version of this manuscript contained a standard quantitative trait model. I think it was a good idea to remove this material, but I still think that this modelling approach (including Fisher’s geometric model) needs to be discussed a little more here. In the introduction, for example, it seems important to state clearly that ``kryptotypes'' (many-to-one genotype-to-phenotype maps) are also a feature of these standard models, and that previous authors have shown how phenotypically identical populations can evolve RI, in some cases via drift alone (e.g., Mani and Clarke 1990; Rosas et al. 2010; Fraisse et al. 2016 and others).

Martin 2014 argued that the results of these models and ``systems-biology'' type models are likely to approximate one another under some conditions, and that could be mentioned.
\end{point}

\reply{
    In the last paragraph of the introduction we discuss exactly this and the corresponding literature, including Rosas et al 2010 and Fraisse et al 2016. \revref
\jss{I don't really think Mani and Clarke says what R1 says it does. It states that mutational order matters. Also I don't thik the introduction is a good place for Martin 2014 -- maybe system and quant gen models have some sort of correspondence, but we don't show it.}
}

@@

\reply{
  The fitness of heterogametic F1s and F2s will always depend on the genetic distance between parents and on the particular gene network. Thus it is not guaranteed that the heterogametic F1s, will always have lower fitnesses than homogametic F1s. This is clear when the genetic distance is zero between parents. The model does show, however, that the particular combination of genes found in F1 heterogametes would only be possible in F2 autosomal crosses. Our model shows that phenotypes generated by networks contained on the autosomes have lower fitnesses in F2s than in F1s, and by extension, lower fitness in the heterogametic sex when networks are contained, at least in part, on the sex chromosomes.

  It is also true that some F2 individuals will perform as well as F1s and F0s -- for instance it is possible to recover an F0-like genotype (and therefore phenotype) after two generations of hybridization. In this model, hybrids have lower fitness because of increased genetic variance. However, there will also be F2 individuals with genes from both parental types -- meaning that there will be more low fitness F2s than low fitness F1s and F0s. 
}

\begin{point}{}
  \hyperlink{rev2:1}{(p. 2, l. 73)} \jss{Note: R2 is using line numbers from the diff}: This paragraph summarises the main body of work performed in this paper. Thus, consider moving it to the end of the Introduction (i.e. switching places with the last paragraph).
\end{point}

\reply{Thank you for this suggestion. We've decided to keep the paragraph order the same though.\jss{Having second thoughts on this... rearranging could be clearer.}
}

### 2021-08-10 kalman_walk_paper.tex
.

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}%\revpoint{1}{5},
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
%\revpoint{1}{1} %\revpoint{1}{3}
% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
% \plr{'structural plasticity' means something specific in neuroscience}

This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
is a consequence of a many-to-one mapping between a system's structure and function,
a concept that has been explored in many fields beyond biology.
For instance, in many contexts mathematical models
are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- %\revpoint{1}{2}
meaning that there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modeled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes
referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

In this paper we use results on mathematical nonidentifiability from\hypertarget{rev2:1}
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}.  %\revpoint{2}{12}
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
%that phenotype does not prescribe a unique genotype\revpoint{2}{3}
that there is often more than one way to do the same thing,
\hypertarget{rev2:2}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. %\revpoint{1}{10}
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. %\revpoint{1}{9}
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$
into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions on the input
(e.g., $|u(t)|$ is integrable) %\revpoint{AE}{3}
we can write the phenotype as %\revpoint{AE}{4}
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

\citet{wagner1994evolution} and others have used a similar discrete-time model %\revpoint{1}{6}
(that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
Our choice of continuous time does not affect the points we make here,
but our restriction to \emph{linear} systems is a stronger assumption
(see the Discussion).

@@

\begin{example}[An oscillator]\label{ex:oscillator}\hypertarget{rev1:1}
  For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}
If the input is an impulse at time zero (a delta function), 
then the phenotype is equal to the impulse response: %\revpoint{AE}{5}
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

@@

One way to find other systems phenotypically equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
are phenotypically equivalent because their impulse response functions are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B} = \bar h(t).
    \end{aligned}
  \end{equation}
These ``changes of coordinates'' are not simply different ways of looking at the same system --
if each dimension of the kryptotype 
corresponds to the concentration of a particular transcription factor,
changing $A$ corresponds to changing the strengths of regulatory interactions.
We will even see below that interactions may change sign. %\revpoint{1}{13}
However, not all phenotypically equivalent systems are of this form:
systems can have identical impulse responses without being coordinate changes of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecules, and thus 
have kryptotypes in
different dimensions altogether.

@@

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any time-invariant\revpoint{2}{5} system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply, in principle, to any system.

@@

Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
gives a change of basis $P$ such that
the transformed system $(PAP^{-1},PB,CP^{-1})$
can be written in block matrix form:
\begin{align*}
       PAP^{-1}
       &=
       \left[ \begin{array}{cccc}
           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
          0 & 0 & 0 & A_{\nro}
       \end{array} \right] ,
\end{align*}
and
\begin{align*}
     PB
     &=
    \left[ \begin{array}{cccc}
         B_{\rno} \\
         B_{\ro} \\
         0 \\
         0 
   \end{array} \right] 
   &
   (CP^{-1})^T
   &=
   \left[ \begin{array}{cccc}
       0 \\
       C_{\ro}^T \\
       0 \\
       C_{\nro}^T
   \end{array} \right] .
\end{align*}
The $n$-dimensional system has been divided into subspaces
of dimensions $n_\rno + n_\ro + n_\nrno + n_\nro = n$,
and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
in the top left corner of $P A P^{-1}$. %\revpoint{1}{15}
The impulse response of the system is given by
\begin{align*}
      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
\end{align*}
and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.

@@

%\revpoint{AE}{7}

>>>>

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
% \plr{'structural plasticity' means something specific in neuroscience}

This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
is a consequence of a many-to-one mapping between a system's structure and function,
a concept that has been explored in many fields beyond biology.
For instance, in many contexts mathematical models
are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- 
meaning that there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modeled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes
referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.

@@

In this paper we use results on mathematical nonidentifiability from\hypertarget{rev2:1}
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}. 
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
%that phenotype does not prescribe a unique genotype
that there is often more than one way to do the same thing,
\hypertarget{rev2:2}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche}, \revpoint{1}{2}
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

We use a model of gene regulatory networks
that describes the temporal dynamics of
a collection of $n$ coregulating molecules
-- such as transcription factors -- as well as external or environmental inputs.
% We model the temporal dynamics of the concentrations of a collection of $n$ coregulating molecules
% within an organism, that may also be affected by temporally varying signals from the environment.
%% Organisms' phenotypes are constructed by gene by gene and gene by environment interactions. 
%%Here we simply define the \emph{phenotype} to be those aspects of the dynamics directly under natural selection
%%-- the \emph{what}, \emph{when}, and \emph{how much}, of an organism's molecules that are physiologically or otherwise relevant to survival.
We write $\kappa(t)$ for the vector of $n$ molecular concentrations at time $t$.
The vector of $m$ ``inputs'' determined exogenously to the system is denoted $u(t)$,
and the vector of $\ell$ ``outputs'' is denoted $\phi(t)$.
The output is merely a linear function of the internal state:
$\phi_i(t) = \sum_j C_{ij} \kappa_j(t)$
for some matrix $C$.
Since $\phi$ is what natural selection acts on, we refer to it as the \emph{phenotype}
(meaning the ``visible'' aspects of the organism),
and in contrast refer to $\kappa$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
Although $\phi$ may depend on all entries of $\kappa$,
it is usually of lower dimension than $\kappa$,
and we tend to think of it as the subset of molecules relevant for survival.
The dynamics are determined by
the matrix of regulatory coefficients, $A$,
a time-varying vector of inputs $u(t)$,
and a matrix $B$ that encodes the effect of each entry of $u$ on the elements of the kryptotype.
%Thus an organism's phenotype $\phi(t)$ -- a vector of molecular concentrations at time $t$ -- is determined both by the structure and organization of a biological system (\emph{e.g.} a gene regulatory network), given by the triple $(A,B,C)$,
%and by its environment $u(t)$.
The rate at which the $i^\text{th}$ concentration changes
is a weighted sum of the concentrations
as well as the input:
\begin{equation}\label{eqn:system}
   \begin{aligned}
    \dot{\kappa}(t) &= A \kappa(t) + B u(t) \\
    \phi(t) &= C \kappa(t) .
  \end{aligned} 
\end{equation}
Furthermore, we always assume that $\kappa(0) = 0$,
so that the kryptotype measures deviations from initial concentrations. 
Here $A$ can be any $n \times n$ matrix, $B$ any $n \times m$, and $C$ any $\ell \times n$ dimensional matrix,
with usually $\ell$ and $m$ less than $n$.
We think of the system as the triple $(A,B,C)$,
which translates (time-varying) $m$-dimensional input $u(t)$
into the $\ell$-dimensional output $\phi(t)$.
Under quite general assumptions on the input
(e.g., $|u(t)|$ is integrable) 
we can write the phenotype as 
  \begin{align}
    \phi(t) = \int_{0}^{t} C e^{(t-s) A} B u(s) ds ,
  \end{align}
which is a convolution of the input $u(t)$ with the system's \emph{impulse response},
which we denote as $h(t) := Ce^{A t}B$.

@@

\citet{wagner1994evolution} and others have used a similar discrete-time model 
(that might be written $\phi_{t+1} = f(A \phi_t)$, where $f$ is a sigmoid).
Our choice of continuous time does not affect the points we make here,
but our restriction to \emph{linear} systems is a stronger assumption
(see the Discussion).

@@

\begin{example}[An oscillator]\label{ex:oscillator}\hypertarget{rev1:1}
  For illustration, we consider an extremely simplified model of oscillating gene transcription,
as for instance is found in cell cycle control or the circadian rhythm.
% Cellular division is governed by many different processes, however it is thought that its rhythm is partially controlled by oscillating gene transcription \citep{orlando2008global}.
There are two genes, 
whose transcript concentrations are given by $\kappa_1(t)$ and $\kappa_2(t)$, 
and gene-2 upregulates gene-1, while gene-1 downregulates gene-2 with equal strength.
Only the dynamics of gene-1 are consequential to the oscillator 
(perhaps the amount of gene-1 activates another downstream gene network). 
Lastly, both genes are equally upregulated by an exogenous signal.
The dynamics of the system are described by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
        \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t) \\
        \phi(t) &= \kappa_{1}(t)   .
    \end{align*}
In matrix form the system regulatory coefficients are given as,
$A \!=\! \left[\begin{smallmatrix} 0 & 1 \\ -1 & 0 \end{smallmatrix}\right]$, 
$B \! =\! \left[\begin{smallmatrix} 1 \\ 1 \end{smallmatrix}\right]$,
and $C \!=\! \left[\begin{smallmatrix} 1 & 0 \end{smallmatrix}\right]$.
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}
If the input is an impulse at time zero (a delta function), 
then the phenotype is equal to the impulse response: 
\begin{align*}
\phi(t) = h(t) = \sin t + \cos t  .
\end{align*}
The system and its dynamics are referred to in Figure \ref{fig:oscillator}. We return to the evolution of such a system below.   
\end{example}

@@

One way to find other systems phenotypically equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
are phenotypically equivalent because their impulse response functions are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B} = \bar h(t).
    \end{aligned}
  \end{equation}
These ``changes of coordinates'' are not simply different ways of looking at the same system --
if each dimension of the kryptotype 
corresponds to the concentration of a particular transcription factor,
changing $A$ corresponds to changing the strengths of regulatory interactions.
We will even see below that interactions may change sign. 
However, not all phenotypically equivalent systems are of this form:
systems can have identical impulse responses without being coordinate changes of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecules, and thus 
have kryptotypes in
different dimensions altogether.

@@

It turns out that in general, there are more degrees of freedom,
except if the system is \emph{minimal} -- meaning, informally, that it uses the smallest possible number of components
to achieve the desired dynamics.
Results in system theory show that any time-invariant \revpoint{2}{5}
system can be realized in a particular minimal dimension
(the dimension of the kryptotype, $n_\text{min}$),
and that any two phenotypically equivalent systems of dimension $n_\text{min}$ are related by a change of coordinates.
Since gene networks can grow or shrink following gene duplications and deletions, 
these additional degrees of freedom can apply, in principle, to any system.

@@

Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
gives a change of basis $P$ such that
the transformed system $(PAP^{-1},PB,CP^{-1})$
can be written in block matrix form:
\begin{align*}
       PAP^{-1}
       &=
       \left[ \begin{array}{cccc}
           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
          0 & 0 & 0 & A_{\nro}
       \end{array} \right] ,
\end{align*}
and
\begin{align*}
     PB
     &=
    \left[ \begin{array}{cccc}
         B_{\rno} \\
         B_{\ro} \\
         0 \\
         0 
   \end{array} \right] 
   &
   (CP^{-1})^T
   &=
   \left[ \begin{array}{cccc}
       0 \\
       C_{\ro}^T \\
       0 \\
       C_{\nro}^T
   \end{array} \right] .
\end{align*}
The $n$-dimensional system has been divided into subspaces
of dimensions $n_\rno + n_\ro + n_\nrno + n_\nro = n$,
and so, for instance, $A_{\rno}$ is the $n_\rno \times n_\rno$ square matrix 
in the top left corner of $P A P^{-1}$. 
The impulse response of the system is given by
\begin{align*}
      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
\end{align*}
and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.

@@
